---
title: "Selective inference across NHANES cycles: lmFScreen example"
author: "Olivia McGough"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Selective inference across NHANES cycles: lmFScreen example}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

*This analysis is meant to mimic the analysis in the paper ["Decreasing Trend of Bone Mineral Density in US Multiethnic Population: Analysis of Continuous NHANES 2005 - 2014"](https://pubmed.ncbi.nlm.nih.gov/30091065/). The results of this analysis can be found in the 2025 paper "Valid F-screening in linear regression" by McGough, Witten, and Kessler (arxiv preprint: [https://arxiv.org/abs/2505.23113](https://arxiv.org/abs/2505.23113)).* 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

In this analysis we use the `lmFScreen` package to compare femur neck bone mineral density (BMD) across National Health and Nutrition Examination Survey (NHANES) cycles for men over age 30. NHANES is a biennial survey that reports a series of health metrics in a large sample of people in the United States. To conduct this analysis, we first perform an F-test of the "overall hypothesis of no change in BMD across cycles. This overall null hypothesis is rejected, so we subsequently conduct post hoc tests for the individual coefficients that correspond to each year/cycle in the data. We adjust for the fact that we would not have conducted the post-hoc tests if we had not rejected the overall null hypothesis.

## Load Required Libraries

```{r}
library(haven)    # For reading NHANES XPT files
library(dplyr)    # For data manipulation
library(lmFScreen) # For selective inference via F-screening
```

## Load and Merge NHANES Data

We load both BMD and demographic data from four NHANES cycles and merge by participant ID (`SEQN`).

```{r}
read_nhanes_cycle <- function(year) {
  BMD <- read_xpt(paste0("DXXFEM_", year, ".xpt"))
  DEMO <- read_xpt(paste0("DEMO_", year, ".xpt"))
  merge(BMD, DEMO, by = "SEQN")
}

merged_2005_2006 <- read_nhanes_cycle("2005_2006")
merged_2007_2008 <- read_nhanes_cycle("2007_2008")
merged_2009_2010 <- read_nhanes_cycle("2009_2010")
merged_2013_2014 <- read_nhanes_cycle("2013_2014")
```

## Clean and Prepare Data

We retain only age, gender, and femur neck BMD, rename the columns, remove missing values, and restrict to men older than 30.

```{r}
prep_bmd_data <- function(df) {
  df <- df[, c("RIDAGEYR", "DXXNKBMD", "RIAGENDR")]
  colnames(df) <- c("Age", "Femur_Neck_BMD", "Gender")
  df <- na.omit(df)
  df %>% filter(Age > 30)
}

bmd_men <- bind_rows(
  prep_bmd_data(merged_2005_2006) %>% filter(Gender == 1) %>% mutate(Year = "2005-2006"),
  prep_bmd_data(merged_2007_2008) %>% filter(Gender == 1) %>% mutate(Year = "2007-2008"),
  prep_bmd_data(merged_2009_2010) %>% filter(Gender == 1) %>% mutate(Year = "2009-2010"),
  prep_bmd_data(merged_2013_2014) %>% filter(Gender == 1) %>% mutate(Year = "2013-2014")
)
```

## Construct Design Matrix

We create indicator variables for each NHANES cycle.

```{r}
bmd_men <- bmd_men %>% mutate(
  Year = factor(Year, levels = c("2005-2006", "2007-2008", "2009-2010", "2013-2014")),
  Year_2005_2006 = as.integer(Year == "2005-2006"),
  Year_2007_2008 = as.integer(Year == "2007-2008"),
  Year_2009_2010 = as.integer(Year == "2009-2010"),
  Year_2013_2014 = as.integer(Year == "2013-2014")
)
```

## Conduct overall F-test

We run an overall F-test to check if there are significant differences in BMD across the years.

```{r}
model <- lm(Femur_Neck_BMD ~ Year + Age, data = bmd_men)
summary(model)
```

The overall F-statistic is significant, so we proceed with inference on the individual coefficients for each year/cycle.

## Prepare for Pairwise Year Comparisons

We now define a function that:

- Constructs a contrast

- Projects out age and the intercept  

- Runs `lmFScreen.fit` on the resulting matrix, which conducts inference on the individual coefficients for each year in the model while accounting for the overall F-test.

```{r}
do_year_comparison <- function(X, test_col, seed_val) {
  set.seed(seed_val)
  n <- nrow(X)
  Y <- bmd_men$Femur_Neck_BMD
  project_out <- cbind(bmd_men$Age)
  svdP <- svd(project_out, nu = nrow(project_out))
  U_perp <- svdP$u[, (sum(svdP$d > .Machine$double.eps) + 1):ncol(svdP$u)]
  X <- t(U_perp) %*% X
  Y <- t(U_perp) %*% Y
  result <- lmFScreen.fit(X, Y, test_cols = test_col, alpha = 0.05, alpha_ov = 0.05)
  summary(result)
}
```

```{r}
do_year_comparison <- function(X, test_col, seed_val) {
  set.seed(seed_val)
  n <- nrow(X)
  Y <- bmd_men$Femur_Neck_BMD
  project_out <- cbind(bmd_men$Age, 1)
  svdP <- svd(project_out, nu = nrow(project_out))
  U_perp <- svdP$u[, (sum(svdP$d > .Machine$double.eps) + 1):ncol(svdP$u)]
  X <- t(U_perp) %*% X
  Y <- t(U_perp) %*% Y
  result <- lmFScreen(Y~X+0, test_cols = test_col, alpha = 0.05, alpha_ov = 0.05)
  summary(result)
}
```

## Run Pairwise Year Comparisons

```{r}
# Compare 2005-2006 vs 2007-2008
do_year_comparison(X=cbind(bmd_men$Year_2005_2006 + bmd_men$Year_2007_2008,
                           bmd_men$Year_2007_2008, bmd_men$Year_2009_2010),
                   test_col = 2, seed_val = 5678)

# Compare 2005-2006 vs 2009-2010
do_year_comparison(X=cbind(bmd_men$Year_2005_2006 + bmd_men$Year_2009_2010,
                           bmd_men$Year_2007_2008, bmd_men$Year_2009_2010),
                   test_col = 3, seed_val = 56910)

# Compare 2005-2006 vs 2013-2014
do_year_comparison(X=cbind(bmd_men$Year_2005_2006 + bmd_men$Year_2013_2014,
                           bmd_men$Year_2009_2010, 
                           bmd_men$Year_2013_2014),
                   test_col = 3, seed_val = 561314)

# Compare 2007-2008 vs 2009-2010
do_year_comparison(X=cbind(bmd_men$Year_2005_2006,
                           bmd_men$Year_2007_2008+bmd_men$Year_2009_2010,
                           bmd_men$Year_2009_2010),
                   test_col = 3, seed_val = 78910)

# Compare 2007-2008 vs 2013-2014
do_year_comparison(X=cbind(bmd_men$Year_2009_2010,
                           bmd_men$Year_2007_2008+bmd_men$Year_2013_2014,
                           bmd_men$Year_2013_2014),
                   test_col = 3, seed_val = 781314)

# Compare 2009-2010 vs 2013-2014
do_year_comparison(X=cbind(bmd_men$Year_2005_2006,
                           bmd_men$Year_2009_2010+bmd_men$Year_2013_2014,
                           bmd_men$Year_2013_2014),
                   test_col = 3, seed_val = 9101314)
```

## Interpretation

Our selective p-values, point estimates, and intervals are valid **conditional on rejecting the overall null hypothesis**, in contrast to standard (naive) linear regression output. However, in this particular case, the selective p-values, estimates, and confidence intervals tend to coincide with their standard counterparts, indicating that selection (i.e. the face that the post hoc tests were conducted conditional on rejecing the overall null hypothesos) did not have a large impact on the analysis. 

