---
title: "Selective inference across NHANES cycles: lmFScreen example"
author: "Olivia McGough"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Selective inference across NHANES cycles: lmFScreen example}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

*This analysis is meant to mimic the analysis in the paper ["Decreasing Trend of Bone Mineral Density in US Multiethnic Population: Analysis of Continuous NHANES 2005 - 2014"](https://pubmed.ncbi.nlm.nih.gov/30091065/). The results of this analysis can be found in ["Valid F-screening in linear regression"](https://arxiv.org/abs/2505.23113).* 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

In this analysis we use the `lmFScreen` package to compare femur neck bone mineral density (BMD) across National Health and Nutrition Examination Survey (NHANES) cycles for men over age 30. This is an annual survey that reports a series of health metrics in a large sample of people in the United States. To conduct this analysis, we first perform an overall F-test to test whether there is a change in BMD across all years. This test is rejected, so we subsequently conduct post-hoc tests for the individual coefficients that correspond to each year/cycle in the data. We adjust for the fact that inference is conditioned on passing a model selection step — in our case, an overall F-test.

## Load Required Libraries

```{r}
library(haven)    # For reading NHANES XPT files
library(dplyr)    # For data manipulation
library(lmFScreen) # For selective inference via F-screening
```

## Load and Merge NHANES Data

We load both BMD and demographic data from four NHANES cycles and merge by participant ID (`SEQN`).

```{r}
read_nhanes_cycle <- function(year) {
  BMD <- read_xpt(paste0("DXXFEM_", year, ".xpt"))
  DEMO <- read_xpt(paste0("DEMO_", year, ".xpt"))
  merge(BMD, DEMO, by = "SEQN")
}

merged_2005_2006 <- read_nhanes_cycle("2005_2006")
merged_2007_2008 <- read_nhanes_cycle("2007_2008")
merged_2009_2010 <- read_nhanes_cycle("2009_2010")
merged_2013_2014 <- read_nhanes_cycle("2013_2014")
```

## Clean and Prepare Data

We retain only age, gender, and femur neck BMD, rename the columns, remove missing values, and restrict to men older than 30.

```{r}
prep_bmd_data <- function(df) {
  df <- df[, c("RIDAGEYR", "DXXNKBMD", "RIAGENDR")]
  colnames(df) <- c("Age", "Femur_Neck_BMD", "Gender")
  df <- na.omit(df)
  df %>% filter(Age > 30)
}

bmd_men <- bind_rows(
  prep_bmd_data(merged_2005_2006) %>% filter(Gender == 1) %>% mutate(Year = "2005-2006"),
  prep_bmd_data(merged_2007_2008) %>% filter(Gender == 1) %>% mutate(Year = "2007-2008"),
  prep_bmd_data(merged_2009_2010) %>% filter(Gender == 1) %>% mutate(Year = "2009-2010"),
  prep_bmd_data(merged_2013_2014) %>% filter(Gender == 1) %>% mutate(Year = "2013-2014")
)
```

## Construct Design Matrix

We create indicator variables for each NHANES cycle.

```{r}
bmd_men <- bmd_men %>% mutate(
  Year = factor(Year, levels = c("2005-2006", "2007-2008", "2009-2010", "2013-2014")),
  Year_2005_2006 = as.integer(Year == "2005-2006"),
  Year_2007_2008 = as.integer(Year == "2007-2008"),
  Year_2009_2010 = as.integer(Year == "2009-2010"),
  Year_2013_2014 = as.integer(Year == "2013-2014")
)
```

## Conduct overall F-test

We run an overall F-test to check if there are significant differences in BMD across the years.

```{r}
model <- lm(Femur_Neck_BMD ~ Year + Age, data = bmd_men)
summary(model)
```

The overall F-statistic is significant, so we proceed with inference on the individual coefficients for each year/cycle.

## Define a Selective Inference Function for Pairwise Year Comparisons

We now define a function that:

- Constructs a contrast

- Projects out age 

- Runs `lmFScreen.fit` on the resulting matrix, which conducts inference on the individual coefficients for each year in the model while accounting for the overall F-test.

```{r}
do_year_comparison <- function(X, test_col, seed_val) {
  set.seed(seed_val)
  n <- nrow(X)
  Y <- bmd_men$Femur_Neck_BMD
  project_out <- cbind(bmd_men$Age)
  svdP <- svd(project_out, nu = nrow(project_out))
  U_perp <- svdP$u[, (sum(svdP$d > .Machine$double.eps) + 1):ncol(svdP$u)]
  X <- t(U_perp) %*% X
  Y <- t(U_perp) %*% Y
  result <- lmFScreen.fit(X, Y, test_cols = test_col, alpha = 0.05, alpha_ov = 0.05)
  summary(result)
}
```

## Run Pairwise Year Comparisons

```{r}
# Compare 2005-2006 vs 2007-2008
do_year_comparison(X=cbind(bmd_men$Year_2005_2006 + bmd_men$Year_2007_2008,
                           bmd_men$Year_2007_2008, bmd_men$Year_2009_2010,
                           bmd_men$Year_2013_2014),
                   test_col = 2, seed_val = 5678)

# Compare 2005-2006 vs 2009-2010
do_year_comparison(X=cbind(bmd_men$Year_2005_2006 + bmd_men$Year_2009_2010,
                           bmd_men$Year_2007_2008, bmd_men$Year_2009_2010,
                           bmd_men$Year_2013_2014),
                   test_col = 3, seed_val = 56910)

# Compare 2005-2006 vs 2013-2014
do_year_comparison(X=cbind(bmd_men$Year_2005_2006 + bmd_men$Year_2013_2014,
                           bmd_men$Year_2007_2008, bmd_men$Year_2009_2010,
                           bmd_men$Year_2013_2014),
                   test_col = 4, seed_val = 561314)

# Compare 2007-2008 vs 2009-2010
do_year_comparison(X=cbind(bmd_men$Year_2005_2006,
                           bmd_men$Year_2007_2008+bmd_men$Year_2009_2010,
                           bmd_men$Year_2009_2010, bmd_men$Year_2013_2014),
                   test_col = 3, seed_val = 78910)

# Compare 2007-2008 vs 2013-2014
do_year_comparison(X=cbind(bmd_men$Year_2005_2006,
                           bmd_men$Year_2007_2008+bmd_men$Year_2013_2014,
                           bmd_men$Year_2009_2010, bmd_men$Year_2013_2014),
                   test_col = 4, seed_val = 781314)

# Compare 2009-2010 vs 2013-2014
do_year_comparison(X=cbind(bmd_men$Year_2005_2006,bmd_men$Year_2007_2008,
                           bmd_men$Year_2009_2010+bmd_men$Year_2013_2014,
                           bmd_men$Year_2013_2014),
                   test_col = 4, seed_val = 9101314)
```

## Interpretation

These analyses provide valid inference **conditional on passing the overall F-test**. The selective p-values, point estimates, and intervals are corrected for the selection step, avoiding overconfident inference that can arise from standard (naive) linear regression output. However, in this particular case, the selective p-values, estimates, and confidence intervals tend to coincide with their standard counterparts, indicating that selection did not have a large impact on the analysis. 

