---
title: "Selective inference across age groups: psel_retro Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selective inference across age groups: psel_retro Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[UTF-8]{inputenc}
---

*Data used in this analysis was extracted from Table 1 of the paper ["Longitudinal Sleep Patterns and Cognitive Impairment in Older Adults"](https://jamanetwork-com.offcampus.lib.washington.edu/journals/jamanetworkopen/fullarticle/2812521).* This table reported means, standard deviations, and sample sizes for four outcome variables (CES-D, Sleep, MMSES, MDRS) across three age groups, along with significant overall p-values for all outcomes (i.e., the overall null hypothesis of no difference between age groups was rejected, for each of the four outcome variables). Therefore, for each outcome variable, we use the `psel_retro()` function from the **lmFScreen** package to compute selective p-values for each pairwise comparison across age groups, conditional on rejection of the overall null hypothesis (i.e., the overall ANOVA). The results of this analysis can be found in the 2025 paper "Valid F-screening in linear regression" by McGough, Witten, and Kessler (arxiv preprint: [https://arxiv.org/abs/2505.23113](https://arxiv.org/abs/2505.23113)).


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lmFScreen)
```

```{r helper-function}
# Function to calculate SE, t-statistic, and p-value for pairwise comparisons
calculate_t_and_p <- function(mean1, mean2, n1, n2, mse, df_within) {
  se <- sqrt(mse * (1/n1 + 1/n2))
  t_stat <- (mean1 - mean2) / se
  p_val <- 2 * pt(-abs(t_stat), df_within)  # two-tailed p-value
  sidak <- 1 - (1 - p_val)^3
  return(list(t_stat = t_stat, p_val = p_val, sidak = sidak))
}
```

---

## CES-D Analysis

```{r}
set.seed(826)
means <- c(9.33, 7.60, 8.79)
sds <- c(7.16, 5.84, 5.68)
ns <- c(150, 449, 227)
n <- sum(ns); p <- 3
overall_mean <- sum(means * ns) / sum(ns)
r_squared <- sum(ns * (means - overall_mean)^2) / (sum(ns * (means - overall_mean)^2) + sum((ns - 1) * sds^2))
mse <- sum((ns - 1) * sds^2) / (n - p)
rse <- sqrt(mse)
df_within <- n - p

results_1_2 <- calculate_t_and_p(means[1], means[2], ns[1], ns[2], mse, df_within)
results_1_3 <- calculate_t_and_p(means[1], means[3], ns[1], ns[3], mse, df_within)
results_2_3 <- calculate_t_and_p(means[2], means[3], ns[2], ns[3], mse, df_within)

cat("\nCES-D Naive p-values (with Sidak correction):\n")
cat("<65 vs 65–84: p =", results_1_2$p_val, ", sidak =", results_1_2$sidak, "\n")
cat("<65 vs ≥85: p =", results_1_3$p_val, ", sidak =", results_1_3$sidak, "\n")
cat("65–84 vs ≥85: p =", results_2_3$p_val, ", sidak =", results_2_3$sidak, "\n")

cat("\nCES-D Selective p-values:\n")
cat("<65 vs 65–84:", psel_retro(n, p, r_squared, rse, (results_1_2$t_stat)), "\n")
cat("<65 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_1_3$t_stat)), "\n")
cat("65–84 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_2_3$t_stat)), "\n")
```

---

## Sleep Analysis

```{r}
set.seed(826)
means <- c(6.65, 6.93, 7.07)
sds <- c(0.87, 0.87, 1.02)
ns <- c(150, 449, 227)
n <- sum(ns); p <- 3
overall_mean <- sum(means * ns) / sum(ns)
r_squared <- sum(ns * (means - overall_mean)^2) / (sum(ns * (means - overall_mean)^2) + sum((ns - 1) * sds^2))
mse <- sum((ns - 1) * sds^2) / (n - p)
rse <- sqrt(mse)
df_within <- n - p

results_1_2 <- calculate_t_and_p(means[1], means[2], ns[1], ns[2], mse, df_within)
results_1_3 <- calculate_t_and_p(means[1], means[3], ns[1], ns[3], mse, df_within)
results_2_3 <- calculate_t_and_p(means[2], means[3], ns[2], ns[3], mse, df_within)

cat("\nSleep Naive p-values (with Sidak correction):\n")
cat("<65 vs 65–84: p =", results_1_2$p_val, ", sidak =", results_1_2$sidak, "\n")
cat("<65 vs ≥85: p =", results_1_3$p_val, ", sidak =", results_1_3$sidak, "\n")
cat("65–84 vs ≥85: p =", results_2_3$p_val, ", sidak =", results_2_3$sidak, "\n")

cat("\nSleep Selective p-values:\n")
cat("<65 vs 65–84:", psel_retro(n, p, r_squared, rse, (results_1_2$t_stat)), "\n")
cat("<65 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_1_3$t_stat)), "\n")
cat("65–84 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_2_3$t_stat)), "\n")
```

---

## MMSES Analysis

```{r}
set.seed(826)
means <- c(29.31, 28.12, 26.14)
sds <- c(1.19, 2.42, 4.15)
ns <- c(150, 449, 227)
n <- sum(ns); p <- 3
overall_mean <- sum(means * ns) / sum(ns)
r_squared <- sum(ns * (means - overall_mean)^2) / (sum(ns * (means - overall_mean)^2) + sum((ns - 1) * sds^2))
mse <- sum((ns - 1) * sds^2) / (n - p)
rse <- sqrt(mse)
df_within <- n - p

results_1_2 <- calculate_t_and_p(means[1], means[2], ns[1], ns[2], mse, df_within)
results_1_3 <- calculate_t_and_p(means[1], means[3], ns[1], ns[3], mse, df_within)
results_2_3 <- calculate_t_and_p(means[2], means[3], ns[2], ns[3], mse, df_within)

cat("\nMMSES Naive p-values (with Sidak correction):\n")
cat("<65 vs 65–84: p =", results_1_2$p_val, ", sidak =", results_1_2$sidak, "\n")
cat("<65 vs ≥85: p =", results_1_3$p_val, ", sidak =", results_1_3$sidak, "\n")
cat("65–84 vs ≥85: p =", results_2_3$p_val, ", sidak =", results_2_3$sidak, "\n")

cat("\nMMSES Selective p-values:\n")
cat("<65 vs 65–84:", psel_retro(n, p, r_squared, rse, results_1_2$t_stat), "\n")
cat("<65 vs ≥85:", psel_retro(n, p, r_squared, rse, results_1_3$t_stat), "\n")
cat("65–84 vs ≥85:", psel_retro(n, p, r_squared, rse, results_2_3$t_stat), "\n")
```

---

## MDRS Analysis

```{r}
set.seed(826)
means <- c(141.5, 139.0, 131.6)
sds <- c(2.57, 5.85, 14.89)
ns <- c(150, 449, 227)
n <- sum(ns); p <- 3
overall_mean <- sum(means * ns) / sum(ns)
r_squared <- sum(ns * (means - overall_mean)^2) / (sum(ns * (means - overall_mean)^2) + sum((ns - 1) * sds^2))
mse <- sum((ns - 1) * sds^2) / (n - p)
rse <- sqrt(mse)
df_within <- n - p

results_1_2 <- calculate_t_and_p(means[1], means[2], ns[1], ns[2], mse, df_within)
results_1_3 <- calculate_t_and_p(means[1], means[3], ns[1], ns[3], mse, df_within)
results_2_3 <- calculate_t_and_p(means[2], means[3], ns[2], ns[3], mse, df_within)

cat("\nMDRS Naive p-values (with Sidak correction):\n")
cat("<65 vs 65–84: p =", results_1_2$p_val, ", sidak =", results_1_2$sidak, "\n")
cat("<65 vs ≥85: p =", results_1_3$p_val, ", sidak =", results_1_3$sidak, "\n")
cat("65–84 vs ≥85: p =", results_2_3$p_val, ", sidak =", results_2_3$sidak, "\n")

cat("\nMDRS Selective p-values:\n")
cat("<65 vs 65–84:", psel_retro(n, p, r_squared, rse, results_1_2$t_stat), "\n")
cat("<65 vs ≥85:", psel_retro(n, p, r_squared, rse, results_1_3$t_stat), "\n")
cat("65–84 vs ≥85:", psel_retro(n, p, r_squared, rse, results_2_3$t_stat), "\n")
```


---

## Summary

In this analysis, we revisited pairwise comparisons across age groups for four outcomes reported in the JAMA Network Open paper ["Longitudinal Sleep Patterns and Cognitive Impairment in Older Adults"](https://jamanetwork-com.offcampus.lib.washington.edu/journals/jamanetworkopen/fullarticle/2812521), by accounting for the fact that these pairwise comparisons would not have been conducted if the overall null hypothesis of no difference across age groups had not been conducted. While some selective p-values differ meaningfully from their standard/naive counterparts (in the case of the CES-D outcome), others remain similar. The Sidak-corrected p-values reported in the original paper align with the Sidak-corrected naive p-values shown here.

By conditioning on rejection of the overall null hypothesis, the selective p-values provide a more principled post-hoc analysis by guarding against inflated Type I error.

