---
title: "Selective Inference on Age Groups"
output: html_document
---

*Data used in this analysis was extracted from Table 1 of the paper ["Longitudinal Sleep Patterns and Cognitive Impairment in Older Adults"](https://jamanetwork-com.offcampus.lib.washington.edu/journals/jamanetworkopen/fullarticle/2812521).* This table reported means, standard deviations, and sample sizes for several variables across age groups. We use the `psel_retro()` function from the **lmFScreen** package to compute selective p-values for each pairwise comparison, conditional on rejection of the overall F-test (i.e., the omnibus ANOVA), since the table reports significant omnibus p-values for all outcomes.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lmFScreen)
```

```{r helper-function}
# Function to calculate SE, t-statistic, and p-value for pairwise comparisons
calculate_t_and_p <- function(mean1, mean2, n1, n2, mse, df_within) {
  se <- sqrt(mse * (1/n1 + 1/n2))
  t_stat <- (mean1 - mean2) / se
  p_val <- 2 * pt(-abs(t_stat), df_within)  # two-tailed p-value
  sidak <- 1 - (1 - p_val)^3
  return(list(t_stat = t_stat, p_val = p_val, sidak = sidak))
}
```

---

## CES-D Analysis

```{r}
means <- c(9.33, 7.60, 8.79)
sds <- c(7.16, 5.84, 5.68)
ns <- c(150, 449, 227)
n <- sum(ns); p <- 3
r_squared <- sum(ns * (means - mean(means))^2) / (sum(ns * (means - mean(means))^2) + sum((ns - 1) * sds^2))
mse <- sum((ns - 1) * sds^2) / (n - p)
rse <- sqrt(mse)
df_within <- n - p

results_1_2 <- calculate_t_and_p(means[1], means[2], ns[1], ns[2], mse, df_within)
results_1_3 <- calculate_t_and_p(means[1], means[3], ns[1], ns[3], mse, df_within)
results_2_3 <- calculate_t_and_p(means[2], means[3], ns[2], ns[3], mse, df_within)

cat("\nCES-D Naive p-values (with Sidak correction):\n")
cat("<65 vs 65–84: p =", results_1_2$p_val, ", sidak =", results_1_2$sidak, "\n")
cat("<65 vs ≥85: p =", results_1_3$p_val, ", sidak =", results_1_3$sidak, "\n")
cat("65–84 vs ≥85: p =", results_2_3$p_val, ", sidak =", results_2_3$sidak, "\n")

cat("\nCES-D Selective p-values:\n")
cat("<65 vs 65–84:", psel_retro(n, p, r_squared, rse, (results_1_2$t_stat)^2), "\n")
cat("<65 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_1_3$t_stat)^2), "\n")
cat("65–84 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_2_3$t_stat)^2), "\n")
```

---

## Sleep Analysis

```{r}
means <- c(6.65, 6.93, 7.07)
sds <- c(0.87, 0.87, 1.02)
ns <- c(150, 449, 227)
n <- sum(ns); p <- 3
r_squared <- sum(ns * (means - mean(means))^2) / (sum(ns * (means - mean(means))^2) + sum((ns - 1) * sds^2))
mse <- sum((ns - 1) * sds^2) / (n - p)
rse <- sqrt(mse)
df_within <- n - p

results_1_2 <- calculate_t_and_p(means[1], means[2], ns[1], ns[2], mse, df_within)
results_1_3 <- calculate_t_and_p(means[1], means[3], ns[1], ns[3], mse, df_within)
results_2_3 <- calculate_t_and_p(means[2], means[3], ns[2], ns[3], mse, df_within)

cat("\nSleep Naive p-values (with Sidak correction):\n")
cat("<65 vs 65–84: p =", results_1_2$p_val, ", sidak =", results_1_2$sidak, "\n")
cat("<65 vs ≥85: p =", results_1_3$p_val, ", sidak =", results_1_3$sidak, "\n")
cat("65–84 vs ≥85: p =", results_2_3$p_val, ", sidak =", results_2_3$sidak, "\n")

cat("\nSleep Selective p-values:\n")
cat("<65 vs 65–84:", psel_retro(n, p, r_squared, rse, (results_1_2$t_stat)^2), "\n")
cat("<65 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_1_3$t_stat)^2), "\n")
cat("65–84 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_2_3$t_stat)^2), "\n")
```

---

## MMSES Analysis

```{r}
means <- c(29.31, 28.12, 26.14)
sds <- c(1.19, 2.42, 4.15)
ns <- c(150, 449, 227)
n <- sum(ns); p <- 3
r_squared <- sum(ns * (means - mean(means))^2) / (sum(ns * (means - mean(means))^2) + sum((ns - 1) * sds^2))
mse <- sum((ns - 1) * sds^2) / (n - p)
rse <- sqrt(mse)
df_within <- n - p

results_1_2 <- calculate_t_and_p(means[1], means[2], ns[1], ns[2], mse, df_within)
results_1_3 <- calculate_t_and_p(means[1], means[3], ns[1], ns[3], mse, df_within)
results_2_3 <- calculate_t_and_p(means[2], means[3], ns[2], ns[3], mse, df_within)

cat("\nMMSES Naive p-values (with Sidak correction):\n")
cat("<65 vs 65–84: p =", results_1_2$p_val, ", sidak =", results_1_2$sidak, "\n")
cat("<65 vs ≥85: p =", results_1_3$p_val, ", sidak =", results_1_3$sidak, "\n")
cat("65–84 vs ≥85: p =", results_2_3$p_val, ", sidak =", results_2_3$sidak, "\n")

cat("\nMMSES Selective p-values:\n")
cat("<65 vs 65–84:", psel_retro(n, p, r_squared, rse, (results_1_2$t_stat)^2), "\n")
cat("<65 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_1_3$t_stat)^2), "\n")
cat("65–84 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_2_3$t_stat)^2), "\n")
```

---

## MDRS Analysis

```{r}
means <- c(141.5, 139.0, 131.6)
sds <- c(2.57, 5.85, 14.89)
ns <- c(150, 449, 227)
n <- sum(ns); p <- 3
r_squared <- sum(ns * (means - mean(means))^2) / (sum(ns * (means - mean(means))^2) + sum((ns - 1) * sds^2))
mse <- sum((ns - 1) * sds^2) / (n - p)
rse <- sqrt(mse)
df_within <- n - p

results_1_2 <- calculate_t_and_p(means[1], means[2], ns[1], ns[2], mse, df_within)
results_1_3 <- calculate_t_and_p(means[1], means[3], ns[1], ns[3], mse, df_within)
results_2_3 <- calculate_t_and_p(means[2], means[3], ns[2], ns[3], mse, df_within)

cat("\nMDRS Naive p-values (with Sidak correction):\n")
cat("<65 vs 65–84: p =", results_1_2$p_val, ", sidak =", results_1_2$sidak, "\n")
cat("<65 vs ≥85: p =", results_1_3$p_val, ", sidak =", results_1_3$sidak, "\n")
cat("65–84 vs ≥85: p =", results_2_3$p_val, ", sidak =", results_2_3$sidak, "\n")

cat("\nMDRS Selective p-values:\n")
cat("<65 vs 65–84:", psel_retro(n, p, r_squared, rse, (results_1_2$t_stat)^2), "\n")
cat("<65 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_1_3$t_stat)^2), "\n")
cat("65–84 vs ≥85:", psel_retro(n, p, r_squared, rse, (results_2_3$t_stat)^2), "\n")
```


---

## Summary

In this analysis, we used selective inference techniques to revisit pairwise comparisons across age groups for four outcomes reported in the JAMA Network Open paper. While some selective p-values differ meaningfully from their naive counterparts (in the case of the CES-D outcome), others remain similar. The Sidak-corrected p-values reported in the original paper align with the Sidak-corrected naive p-values shown here.

By conditioning on rejection of the overall F-test, the selective p-values provide a more principled post-hoc analysis while guarding against inflated Type I error.

